---
title: "Module 5 Report"
author: "vRoom-vRoom"
date: "`r Sys.Date()`"
output: html_document
bibliography: Biol3140.bib
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
dat.f <- list.files(pattern = "chair|active.csv") #find the files

dat.l <- list() #make an empty list

for(i in dat.f){
  met.dat<- unlist(strsplit(i,"_")) #split of the filename string
  who <- met.dat[1] #store the first element
  activity <- gsub(".csv","",met.dat[2]) #store second element, dropping CSV with gsub()
  dat.l[[i]]<- read_csv(i)%>%
    mutate(who=who,activity=activity) #read, then store data with additional columns of metada in list as position unique to i
  
}
dat <- do.call(rbind,dat.l) #combine the list into table

```

## Introduction

A study conducted by @lim2008human discusses the importance of thermoregulation in protecting the body from the consequences related to the over heating of the body, claiming that these consequences start arising at 39.5°C and that without thermoregulatoin the body could reach 42.0°C during strenuous exercise. This led us to pose the question does body temperature still vary in human subjects according to activity level?

## Methods

The four participants in this study sat in a chair for five minutes and then had their temperature recorded for 100 seconds. They then did 5 minutes of jumping jacks and immediately got their temperature recorded for another 100 seconds.

The temperature was recorded using a DHT11 sensor connected to an Arduino board. The DHT11 sensor was placed in a small plastic bag and in the participant's armpit for 5 minutes so that the sensor warmed up to body temperature and the data was then recorded 100 seconds.

## Results

```{r, echo=FALSE}

dat%>%
  ggplot(aes(x=activity,y=degC))+geom_boxplot()+facet_grid(.~who)  #plot to see what we get

```

## Discussion

## Author Contributions

## References