---
title: "Module 5 Report"
author: "vRoom-vRoom"
date: "`r Sys.Date()`"
output: html_document
bibliography: Biol3140.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, echo=FALSE}

```

## Introduction

A study conducted by @lim2008human discusses the importance of thermoregulation in protecting the body from the consequences related to the over heating of the body, claiming that these consequences start arising at 39.5°C and that without thermoregulatoin the body could reach 42.0°C during strenuous exercise. This led us to pose the question does body temperature still vary in human subjects according to activity level?

## Methods

The four participants in this study sat in a chair for five minutes and then had their temperature recorded for 100 seconds and then did 5 minutes of jumping jacks and immediately got their temperature recorded for another 100 seconds.

The temperature was recorded using a DHT11 sensor connected to an Arduino board. The DHT11 sensor was placed in a small plastic bag and in the particitant's armpit during the time of data collection.

## Results

## Discussion

## Author Contributions

## References